\TitlePt

### 2016-03-24 paper-moho.tex
Finish foundation of intro text
Need to refine the wording and expand the text a bit
Add submitted version of the moho paper

\chapter{Fast non-linear gravity inversion in spherical coordinates with
         application to the South American Moho}

@@

Estimating the relief of the Moho from gravity data is a computationally
intensive non-linear inverse problem.
What is more, the modeling must take the Earths curvature into account when
the study area is of regional scale or greater.
We present a regularized non-linear gravity inversion method that has a low
computational footprint and employs a spherical Earth approximation.
To achieve this, we combine the highly efficient Bott's method with
smoothness regularization and a discretization of the anomalous Moho into
tesseroids (spherical prisms).
The computational efficiency of our method is attained by harnessing the
fact that all matrices involved are sparse.
The inversion results are controlled by three hyper-parameters: the
regularization parameter, the anomalous Moho density-contrast, and the
reference Moho depth.
We estimate the regularization parameter using the method of hold-out
cross-validation.
Additionally, we estimate the density-contrast and the reference depth
using knowledge of the Moho depth at certain points.
We apply the proposed method to estimate the Moho depth for the South
American continent using satellite gravity data and seismological data.
The final Moho model is in accordance with previous gravity-derived models
and seismological data.
The misfit to the gravity and seismological data is worse in the Andes and
best in oceanic areas, central Brazil and Patagonia, and along the Atlantic
coast.
Similarly to previous results, the model suggests a thinner crust of 30-35
km under the Andean foreland basins.
Discrepancies with the seismological data are greatest in the Guiana
shield, the central Solimões and Amazon basins, the Paraná basin, and the
Borborema province.
These differences suggest the existence of crustal or mantle density
anomalies that were unaccounted for during gravity data processing.

@@

The Mohorovičić discontinuity (or Moho) that marks the transition from the
crust to the mantle, is studied almost exclusively through indirect geophysical
methods.
The two main methods used to estimate the depth of the Moho are seismology,
with both natural and controlled sources, and gravimetry.
With the advent of satellite gravimetry missions like GRACE and GOCE,
gravity derived crustal models can be produced in regional or global scales
\citep[e.g. ][]{reguzzoni2013,vandermeijde2013,vandermeijde2015}.
New spherical harmonic gravity models that use these satellite observation,
like GOCO5S \citep{mayer-guerr2015}, provide almost homogeneous data coverage
in difficult to access regions traditionally poor in terrestrial data.
An example is South America, where seismologic and terrestrial gravity data
are traditionally concentrated around large urban centers and coastal areas.

Estimating Moho depth from gravity data is a non-linear inverse problem.
One can generalize this problem to estimating the relief of an interface,
such as the basement of a sedimentary basin or the relief of the anomalous
Moho.
Several methods have been developed over the years to solve this inverse
problem, for example
\citet{bott1960, barbosa1999b, barbosa1999a, barnes2012, leao1996, martins2010,
martins2011, oldenburg1974, reguzzoni2013, santos2015, silva2006, silva2014},
to name a few.
Solving the inverse problem is computationally demanding because it requires
the construction of large dense matrices and the solution of large linear
systems.
As a result, some authors search for ways to increase the computational
efficiency of this class of inverse problem.
\citet{bott1960} proposed a method based on iteratively applying corrections to
a starting estimate based on the inversion residuals.
The algorithm is fast because it bypasses the construction and solution of
linear systems and only involves forward modeling.
\citet{oldenburg1974} showed that the fast FFT-based forward modeling of
\citet{parker1973} could be rearranged to estimate the relief.
\citet{barnes2012} use a form of adaptive discretization to compute the
Jacobian, or sensitivity, matrix.
For each data point, the discretization will be progressively coarser
the further way from the point.
This reduces the matrix and, consequently, the linear systems to a sparse form
that can be solved efficiently.
Recently, \citet{silva2014} extended and generalized the original method of
\citet{bott1960} and \citet{santos2015} used this extension to estimate a
basement relief with sharp boundaries.

Most non-linear gravity inversion methods discretize the relief of the
interface into juxtaposed right-rectangular prisms with a known density
contrast.
The inverse problem is then to estimate the thickness of each prism from the
gravity data.
The use of rectangular prisms implies a planar Earth approximation and may not
be adequate for continental and global scale studies.
In such cases, a spherical Earth approximation is preferred.
\citet{wieczorek1998} developed a spherical harmonic equivalent of the
Parker-Oldenburg FFT algorithm and applied it to estimate the crustal structure
of the Moon.
\citet{reguzzoni2013} use a spherical approximation to estimate the global
Moho relief from GOCE satellite gravity data.
Conversely, one could adapt one of the methods developed for
right-rectangular prisms to use tesseroids (spherical prisms) instead.
One of the difficulties of this approach is that the forward problem for a
tesseroid must be solved numerically.
Two alternatives proposed in the literature to the numerical solution are
Taylor series expansion \citep{heck2007, grombein2013}
and the Gauss-Legendre Quadrature
\citep{asgharzadeh2007}.
Numerical experiments by \citet{wild-pfeiffer2008} suggest that the
Gauss-Legendre Quadrature (GLQ) offers superior results.
However, the GLQ suffers from numerical instability when the computation point
is close to the tesseroid \citep{asgharzadeh2007}.
To overcome the numerical instability, \citet{li2011} proposed an adaptive
discretization algorithm which was later improved upon by \citet{uieda2016}.

In any gravity inversion for the relief of an interface, two hyper-parameters
control the inversion results: the density-contrast between the two mediums
and the reference level around which the interface undulates.
The reference level is the depth of the Normal Earth Moho in the case of the
anomalous Moho.
For regularized inversions, an additional hyper-parameter is the regularization
parameter that balances data-misfit and regularization.
The two most commonly used methods for estimating the regularization parameter
are the L-curve criterion and Generalized Cross Validation (GCV).
\citet{farquharson2004} provide for a thorough comparison of both methods.
Estimating the density-contrast in a sedimentary basin context has been tackled
by \citet{silva2006} and \citet{martins2010} when the basement depth is known
at a few points.
To the authors knowledge no attempt has been made to estimate the reference
level.

We present a non-linear gravity inversion to estimate the Moho depth in
a spherical Earth approximation.
Our method is based on the \citet{silva2014} Gauss-Newton formulation of the
method of \citet{bott1960}.
We use tesseroids to discretize the anomalous Moho and the adaptive
discretization algorithm of \citet{uieda2016} for the forward modeling.
The stability of the inversion is achieved through smoothness regularization.
In order to maintain the computational efficiency of Bott's method,
we exploit the sparse nature of all matrices involved in the computations.
We employ a variant of GCV known as hold-out cross-validation \citep{kim2009}
to estimate the regularization parameter.
Additionally, we estimate the density-contrast and reference level
simultaneously in a second cross-validation.
Similarly to \citet{silva2006} and \citet{martins2010}, this cross-validation
procedure uses knowledge of the Moho depth at certain points.
Finally, we apply the proposed method to estimate the Moho depth for South
America using gravity data from the GOCO5S model \citep{mayer-guerr2015} and
the seismological data of \citet{assumpcao2013a}.

@@

In potential field methods,
we must isolate the target anomalous density distribution prior to modeling and
inversion.
In our case, the target is the relief of the real Moho undulating around a
reference Moho.
We do this by removing all other effects from the gravity observations.
The first correction is to remove the
scalar gravity of an ellipsoidal reference Earth (the Normal Earth),
hereafter denoted as $\gamma$.
This effect is calculated on the same point $P$ where
the gravity observation was made
(Fig~\ref{fig:p3-anomalysketch}a-b).
$\gamma(P)$ is calculated using
the closed-form solution presented by \citet{li2001a}.
The difference between the observed gravity at point P ($g(P)$)
and Normal gravity at the same point
is known as the gravity disturbance,

\begin{equation}
    \delta(P) = g(P) - \gamma(P).
    \label{eq:p3-disturbance}
\end{equation}

The disturbance contains only the gravitational effects of density
distributions that are anomalous with respect to the Normal Earth
(see Fig.~\ref{fig:p3-anomalysketch}c).
This includes all masses above the surface of the ellipsoid (the topography),
the mass deficiency of the oceans,
the mass deficiency of sedimentary basins,
crustal sources (e.g., igneous intrusions, lateral density changes, etc),
heterogeneities below the upper mantle,
and the effect of the difference between the real Moho
topography and the Moho of the Normal Earth.

In order to invert for the anomalous Moho relief,
we must first isolate its gravitational attraction.
Thus, all other effects
must be either removed or assumed negligible.
Here, we will remove the effect of the topography and oceans
in order to obtain the full Bouguer disturbance
(Fig~\ref{fig:p3-anomalysketch}d),

\begin{equation}
    \delta_{bg}(P) = \delta(P) - g_{topo}(P).
    \label{eq:p3-bouguer}
\end{equation}

\noindent
We will remove the effect of sedimentary basins
but assume that the effects of
other crustal and mantle sources are negligible.
Thus, the only effect left will be that of the anomalous Moho relief
(Fig~\ref{fig:p3-anomalysketch}e).
The gravitational attraction of the topography, oceans, and basins are
calculated in a spherical Earth approximation by forward modeling using
tesseroids (Fig.~\ref{fig:p3-tesseroid}).
The tesseroid effects are calculated numerically using
Gauss-Legendre Quadrature (GLQ) integration \citep{asgharzadeh2007}.
The accuracy of the GLQ integration is improved by the adaptive discretization
scheme of \citet{uieda2016}.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{figures/paper-moho/problem-concept}
    \caption{
        Sketch of the stages in gravity data correction and
        the discretization of the anomalous Moho relief using tesseroids.
        (a) The Earth and the measured gravity at point P ($g(P)$).
        (b) The Normal Earth and the calculated normal gravity at point P
        ($\gamma(P)$). $z_{ref}$ is the depth of the Normal Earth Moho.
        (c) The gravity disturbance ($\delta(P)$) and
        the corresponding density anomalies after removal of the normal gravity:
        topography, oceans, crustal heterogeneities, and the anomalous Moho.
        (d) The Bouguer disturbance ($\delta_{bg}(P)$) after topographic
        correction and the remaining density anomalies.
        (e) All density anomalies save the anomalous Moho are assumed to have
        been removed before inversion.
        (f) The discretization of the anomalous Moho in tesseroids. Grey
        tesseroids will have a negative density contrast while red tesseroids
        will have a positive one.
    }
    \label{fig:p3-anomalysketch}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/paper-moho/tesseroid-coord-sys}
    \caption{Sketch of a tesseroid (spherical prism) in a geocentric coordinate
        system (X, Y, Z).
        Observations are made at point P with respect to it's local
        North-oriented coordinate system (x, y, z).
        After \citet{uieda2015}.
    }
    \label{fig:p3-tesseroid}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametrization}

We parameterize the forward problem by discretizing the anomalous Moho
into a grid of $M_{lon} \times M_{lat} = M$ juxtaposed tesseroids
(Fig~\ref{fig:p3-anomalysketch}f).
The true (real Earth) Moho varies in depth
with respect to the Moho of the Normal Earth.
Hereafter we will refer to the depth of the Normal Earth Moho as $z_{ref}$
(see Fig.~\ref{fig:p3-anomalysketch}b).
In cases where the true Moho is above $z_{ref}$,
the top of the $k$th tesseroid is the Moho depth $z_{k}$,
the bottom is $z_{ref}$, and the density-contrast ($\Delta\rho$) is positive
(red tesseroids in Fig~\ref{fig:p3-anomalysketch}f).
If the Moho is below $z_{ref}$, the top of the tesseroid is $z_{ref}$,
the bottom is $z_k$, and $\Delta\rho$ is negative
(grey tesseroids in Fig~\ref{fig:p3-anomalysketch}f).

@@

\begin{equation}
    d_i = f_i(\mathbf{p}),
    \label{eq:p3-forward}
\end{equation}

@@

We wish to estimate the parameter vector $\mathbf{p}$ from a set of observed
gravity anomaly data $\mathbf{d}^o$.
The least-squares estimate is the one that minimizes the data-misfit function

\begin{equation}
    \phi(\mathbf{p}) =
    [\mathbf{d}^o - \mathbf{d}(\mathbf{p})]^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p})].
    \label{eq:p3-data-misfit}
\end{equation}

Function $\phi(\mathbf{p})$ is non-linear with respect to $\mathbf{p}$.
Thus, we can determine its minimum using gradient-based
iterative optimization
methods like Gauss-Newton or Steepest Descent.
Such methods start from an initial estimate $\mathbf{p}^0$ and iteratively
update the estimate until a minimum is reached.

For the Gauss-Newton method,
the update at the $k$th iteration,
$\mathbf{\Delta p} = \mathbf{p}^{k+1} - \mathbf{p}^k$,
is the solution of the linear system

\begin{equation}
    \mathbf{H}^k\mathbf{\Delta p} = -\mathbf{\nabla\phi}^k,
    \label{eq:p3-gaussnewton}
\end{equation}

@@

The Steepest Descent method uses only the gradient direction
to update the initial estimate \citep{kelley1987}.
The update at the $k$th iteration is achieved by equating the Hessian
in Eq.~\ref{eq:p3-gaussnewton} to the identity matrix,

\begin{equation}
    \mathbf{\Delta p} = -\mathbf{\nabla\phi}^k.
    \label{eq:p3-steepest}
\end{equation}

\noindent
Thus, it does not require the computation and storage of the Hessian matrix
nor the solution of linear systems.
However, the Steepest Descent method has poor convergence when the
current solution is close to the minimum of the goal function
\citep{kelley1987}.

@@

\begin{equation}
    \mathbf{\nabla\phi}^k = -2\mathbf{A}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)],
    \label{eq:p3-gradient}
\end{equation}

@@

\begin{equation}
    \mathbf{H}^k \approx 2\mathbf{A}^T\mathbf{A},
    \label{eq:p3-hessian}
\end{equation}

\noindent in which
$\mathbf{A}$ is the Jacobian or sensitivity matrix,

\begin{equation}
    A_{ij} = \dfrac{\partial f_i}{\partial p_j}(\mathbf{p^k}).
    \label{eq:p3-jacobian}
\end{equation}

@@

Non-linear inversions for the relief of an interface (like the Moho)
are ill-posed and require additional constraints in the form of
regularization \citep{silva2001}.
A common approach is to use the first-order Tikhonov regularization to impose
smoothness on the solution.
The cost function for smoothness regularization is given by

\begin{equation}
    \theta(\mathbf{p}) = \mathbf{p}^T\mathbf{R}^T\mathbf{R}\mathbf{p},
    \label{eq:p3-regul}
\end{equation}

\noindent where $\mathbf{R}$ is an $L \times M$ finite-difference matrix
representing the $L$ first-order differences between adjacent tesseroids.

The solution $\hat{\mathbf{p}}$ to the regularized inverse problem is the one that
minimizes the goal function

\begin{equation}
    \Gamma(\mathbf{p}) = \phi(\mathbf{p}) + \mu\theta(\mathbf{p}),
    \label{eq:p3-goalfunction}
\end{equation}

\noindent
in which $\mu$ is the regularization parameter that controls the balance
between fitting the observed data and obeying the smoothness constraint.

The goal function $\Gamma(\mathbf{p})$ is also non-linear with respect to
$\mathbf{p}$ and can be minimized using the Gauss-Newton or Steepest Descent
methods.
The gradient vector and Hessian matrix of the goal function are, respectively,

\begin{equation}
    \mathbf{\nabla\Gamma}^k =
        -2\mathbf{A}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)] +
        2\mu\mathbf{R}^T\mathbf{R}\mathbf{p}^k,
    \label{eq:p3-gradient-regul}
\end{equation}

@@

\begin{equation}
    \mathbf{H}^k = 2\mathbf{A}^T\mathbf{A} + 2\mu\mathbf{R}^T\mathbf{R}.
    \label{eq:p3-hessian-regul}
\end{equation}

\noindent The parameter updates for the regularized Gauss-Newton and Steepest
Descent methods, respectively, then become

\begin{equation}
    \left[\mathbf{A}^T\mathbf{A} + \mu\mathbf{R}^T\mathbf{R}\right]
    \mathbf{\Delta p} =
        \mathbf{A}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)] -
        \mu\mathbf{R}^T\mathbf{R}\mathbf{p}^k,
    \label{eq:p3-gaussnewton-regul}
\end{equation}

\begin{equation}
    \mathbf{\Delta p} =
        \mathbf{A}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)] -
        \mu\mathbf{R}^T\mathbf{R}\mathbf{p}^k,
    \label{eq:p3-steepest-regul}
\end{equation}

Producing the regularized solution using the above equations is computationally
costly because of two main factors:
(1) the evaluation and storage of the dense $N \times M$ Jacobian matrix
$\mathbf{A}$
and (2) the solution of the resulting $M \times M$ equation system
(not required for Steepest Descent).
In practice, the derivatives in the Jacobian (Eq.~\ref{eq:p3-jacobian})
are often calculated through a first-order finite-difference approximation.
Thus, evaluating $\mathbf{A}$ requires $2\times M \times N$ forward modeling
operations for each iteration of the gradient descent algorithm.
These computations are performed for each iteration of the optimization.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software implementation}\label{sec:p3-software}

>>>>

\chapter{Fast non-linear gravity inversion in spherical coordinates with
         application to the South American Moho}
\label{chap:moho}

@@

Estimating the relief of the Moho from gravity data is a computationally
intensive non-linear inverse problem.  What is more, the modeling must take the
Earths curvature into account when the study area is of regional scale or
greater.  We present a regularized non-linear gravity inversion method that has
a low computational footprint and employs a spherical Earth approximation.  To
achieve this, we combine the highly efficient Bott's method with smoothness
regularization and a discretization of the anomalous Moho into tesseroids
(spherical prisms).  The computational efficiency of our method is attained by
harnessing the fact that all matrices involved are sparse.  The inversion
results are controlled by three hyper-parameters: the regularization parameter,
the anomalous Moho density-contrast, and the reference Moho depth.  We estimate
the regularization parameter using the method of hold-out cross-validation.
Additionally, we estimate the density-contrast and the reference depth using
knowledge of the Moho depth at certain points.  We apply the proposed method to
estimate the Moho depth for the South American continent using satellite
gravity data and seismological data.  The final Moho model is in accordance
with previous gravity-derived models and seismological data.  The misfit to the
gravity and seismological data is worse in the Andes and best in oceanic areas,
central Brazil and Patagonia, and along the Atlantic coast.  Similarly to
previous results, the model suggests a thinner crust of 30-35 km under the
Andean foreland basins.  Discrepancies with the seismological data are greatest
in the Guyana Shield, the central Solimões and Amazonas Basins, the Paraná
Basin, and the Borborema province.  These differences suggest the existence of
crustal or mantle density anomalies that were unaccounted for during gravity
data processing.

@@

The Mohorovičić discontinuity (or Moho) that marks the transition from the
crust to the mantle, is studied almost exclusively through indirect geophysical
methods.
The two main geophysical methods used to estimate the depth of the Moho are
seismology, with both natural and controlled sources, and gravimetry.
With the advent of satellite gravimetry missions like GRACE and GOCE,
gravity-derived crustal models can be produced in regional or global scales
\citep[e.g. ][]{reguzzoni2013,vandermeijde2013,vandermeijde2015}.
New spherical harmonic gravity models that use these satellite observation,
like GOCO5S \citep{mayer-guerr2015}, provide almost homogeneous data coverage
in difficult to access regions traditionally poor in terrestrial data.
An example is South America, where seismologic and terrestrial gravity data
are traditionally concentrated around urban centers and coastal areas,
resulting in large areas (e.g., forests and mountains) devoid of data.

Estimating Moho depth from gravity data is a non-linear inverse problem.
One can generalize this problem of estimating the depths of an interface
separating two media,
such as the sediment-basement interface of a sedimentary basin or the
crust-mantle interface (Moho).
Several methods have been developed over the years to solve this inverse
problem, for example
\citet{bott1960, barbosa1999b, barbosa1999a, barnes2012, leao1996, martins2010,
martins2011, oldenburg1974, reguzzoni2013, santos2015, silva2006, silva2014},
to name a few.
Solving the inverse problem is computationally demanding because it requires
the construction of large dense matrices and the solution of large linear
systems.
As a result, some authors search for ways to increase the computational
efficiency of this class of inverse problem.
\citet{bott1960} proposed a method based on iteratively applying corrections to
a starting estimate based on the inversion residuals.
The algorithm is fast because it bypasses the construction and solution of
linear systems and only involves forward modeling.
\citet{oldenburg1974} showed that the fast FFT-based forward modeling of
\citet{parker1973} could be rearranged to estimate the relief.
\citet{barnes2012} use a form of adaptive discretization to compute the
Jacobian, or sensitivity, matrix.
For each data point, the discretization will be progressively coarser
the further way from the point.
This reduces the matrix and, consequently, the linear systems to a sparse form
that can be solved efficiently.
Recently, \citet{silva2014} extended and generalized the original method of
\citet{bott1960} and \citet{santos2015} used this extension to estimate a
basement relief with sharp boundaries.

A spherical Earth approximation is preferred when estimating the Moho depth
from gravity data in continental and global scale studies.
\citet{wieczorek1998} developed a spherical harmonic equivalent of the
Parker-Oldenburg FFT algorithm and applied it to estimate the crustal structure
of the Moon.
\citet{reguzzoni2013} use a spherical Earth approximation to estimate the
global Moho relief using data from the GOCE satellite mission.
Another approach is to use
non-spectral (space domain) gravity inversion methods.
Many such methods were developed for estimating the basement relief of a
sedimentary basin
\citep[e.g., ][]{barbosa1999a, barbosa1999b, martins2010, martins2011, sun2014}.
These methods approximate the sedimentary pack by a set of juxtaposed
right-rectangular prisms.
The top of the prisms coincide with the Earth's surface and the prisms'
thicknesses represent the depths to the basement and are the parameters to be
estimated in the inversion.
The use of rectangular prisms implies a planar Earth approximation and may not
be adequate for depth-to-Moho estimates in a continental-scale study.
A straightforward way to circumvent this hindrance is to adapt one of the
methods developed for rectangular prisms to use tesseroids (spherical prisms).
One of the difficulties of this approach is that the forward problem for a
tesseroid must be solved numerically.
Two alternatives proposed in the literature to the numerical solution are
Taylor series expansion \citep{heck2007, grombein2013}
and the Gauss-Legendre Quadrature
\citep{asgharzadeh2007}.
Numerical experiments by \citet{wild-pfeiffer2008} suggest that the
Gauss-Legendre Quadrature (GLQ) offers superior results.
However, the GLQ suffers from numerical instability when the computation point
is close to the tesseroid \citep{asgharzadeh2007}.
To overcome the numerical instability, \citet{li2011} proposed an adaptive
discretization algorithm which was later improved upon by \citet{uieda2016}.

In any gravity inversion for estimating the relief of an interface, two
hyper-parameters control the inversion results: the density-contrast between
the two media and the reference level around which the interface undulates.
The reference level is the constant depth of the Normal Earth Moho in the case
of the anomalous Moho.
For regularized inversions, an additional hyper-parameter is the regularization
parameter that balances the relative importance between the data-misfit measure
and the regularizing function.
The two most commonly used methods for estimating the regularization parameter
are the L-curve criterion and Generalized Cross Validation (GCV).
\citet{farquharson2004} provide for a thorough comparison of both methods.
Estimating the density-contrast in a sedimentary basin context has been tackled
by \citet{silva2006} and \citet{martins2010} when the basement depth is known
at a few points.
To the authors' knowledge no attempt has been made to estimate the reference
level.

We present a non-linear gravity inversion to estimate the Moho depth in
a spherical Earth approximation.
Our method is based on the \citet{silva2014} Gauss-Newton formulation of the
method of \citet{bott1960}.
We use tesseroids to discretize the anomalous Moho and the adaptive
discretization algorithm of \citet{uieda2016} for the forward modeling.
The stability of the inversion is achieved through smoothness regularization.
In order to maintain the computational efficiency of Bott's method,
we exploit the sparse nature of all matrices involved in the computations.
We employ a variant of GCV known as hold-out cross-validation \citep{kim2009}
to estimate the regularization parameter.
Additionally, we estimate the density-contrast and reference level
simultaneously in a second cross-validation.
Similarly to \citet{silva2006} and \citet{martins2010}, this cross-validation
procedure uses knowledge of the Moho depths at certain points.
Finally, we apply the proposed method to estimate the Moho depth for South
America using gravity data from the GOCO5S model \citep{mayer-guerr2015} and
the seismological data of \citet{assumpcao2013a}.

@@

In potential field methods, we must isolate the target anomalous density
distribution before modeling and inversion.
In our case, the target is the relief of the real Moho undulating around a
reference Moho.
We do this by removing all other effects from the gravity observations.
The first correction is to remove the
scalar gravity of an ellipsoidal reference Earth (the Normal Earth),
hereafter denoted as $\gamma$.
This effect is calculated on the same point $P$ where
the gravity observation was made
(Fig~\ref{fig:moho-anomalysketch}a-b).
$\gamma(P)$ is calculated using
the closed-form solution presented by \citet{li2001a}.
The difference between the observed gravity at point P ($g(P)$)
and Normal gravity at the same point
is known as the gravity disturbance,

\begin{equation}
    \delta(P) = g(P) - \gamma(P).
    \label{eq:moho-disturbance}
\end{equation}

The disturbance contains only the gravitational effects of density
distributions that are anomalous with respect to the Normal Earth
(see Fig.~\ref{fig:moho-anomalysketch}c).
This includes all masses above the surface of the ellipsoid (the topography),
the mass deficiency of the oceans,
the mass deficiency of sedimentary basins,
crustal sources (e.g., igneous intrusions, lateral density changes, etc),
heterogeneities below the upper mantle,
and the effect of the difference between the real Moho
topography and the Moho of the Normal Earth.

To estimate the anomalous Moho relief from gravity data,
we must first isolate its gravitational attraction.
Thus, all other gravitational effects
must be either removed or assumed negligible.
Here, we will remove the gravitational effect produced by the known topography
and ocean masses to obtain the full Bouguer disturbance
(Fig~\ref{fig:moho-anomalysketch}d),

\begin{equation}
    \delta_{bg}(P) = \delta(P) - g_{topo}(P).
    \label{eq:moho-bouguer}
\end{equation}

\noindent
We will also remove the gravitational effect of know sedimentary basins
but assume that the effects of
other crustal and mantle sources are negligible.
Thus, the only effect left will be that of the anomalous Moho relief
(Fig~\ref{fig:moho-anomalysketch}e).
The gravitational attraction of the topography, oceans, and basins are
calculated in a spherical Earth approximation by forward modeling using
tesseroids (Fig.~\ref{fig:moho-tesseroid}).
The tesseroid effects are calculated numerically using
Gauss-Legendre Quadrature (GLQ) integration \citep{asgharzadeh2007}.
The accuracy of the GLQ integration is improved by the adaptive discretization
scheme of \citet{uieda2016}.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/paper-moho/problem-concept}
    \caption{
        Sketch of the stages in gravity data correction and
        the discretization of the anomalous Moho relief using tesseroids.
        (a) The Earth and the measured gravity at point P ($g(P)$).
        (b) The Normal Earth and the calculated normal gravity at point P
        ($\gamma(P)$). $z_{ref}$ is the depth of the Normal Earth Moho.
        (c) The gravity disturbance ($\delta(P)$) and the corresponding density
        anomalies after removal of the normal gravity: topography, oceans,
        crustal and mantle heterogeneities, and the anomalous Moho.
        (d) The Bouguer disturbance ($\delta_{bg}(P)$) after topographic
        correction and the remaining density anomalies.
        (e) All density anomalies save the anomalous Moho are assumed to have
        been removed before inversion.
        (f) The discretization of the anomalous Moho in tesseroids. Grey
        tesseroids will have a negative density contrast while red tesseroids
        will have a positive one.
    }
    \label{fig:moho-anomalysketch}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/paper-moho/tesseroid-coord-sys}
    \caption{Sketch of a tesseroid (spherical prism) in a geocentric coordinate
        system (X, Y, Z).
        Observations are made at point P with respect to it's local
        North-oriented coordinate system (x, y, z).
        After \citet{uieda2015}.
    }
    \label{fig:moho-tesseroid}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametrization and the forward problem}

We parameterize the forward problem by discretizing the anomalous Moho
into a grid of $M_{lon} \times M_{lat} = M$ juxtaposed tesseroids
(Fig~\ref{fig:moho-anomalysketch}f).
The true (real Earth) Moho varies in depth
with respect to the Moho of the Normal Earth.
Hereafter we will refer to the depth of the Normal Earth Moho as $z_{ref}$
(see Fig.~\ref{fig:moho-anomalysketch}b).
If the true Moho is above $z_{ref}$,
the top of the $k$th tesseroid is the Moho depth $z_{k}$,
the bottom is $z_{ref}$, and the density-contrast ($\Delta\rho$) is positive
(red tesseroids in Fig~\ref{fig:moho-anomalysketch}f).
If the Moho is below $z_{ref}$, the top of the tesseroid is $z_{ref}$,
the bottom is $z_k$, and $\Delta\rho$ is negative
(grey tesseroids in Fig~\ref{fig:moho-anomalysketch}f).

@@

\begin{equation}
    d_i = f_i(\mathbf{p}),
    \label{eq:moho-forward}
\end{equation}

@@

We wish to estimate the parameter vector $\mathbf{p}$ from a set of observed
gravity data $\mathbf{d}^o$.
The least-squares estimate is the one that minimizes the data-misfit function

\begin{equation}
    \phi(\mathbf{p}) =
        [\mathbf{d}^o - \mathbf{d}(\mathbf{p})]^T
        [\mathbf{d}^o - \mathbf{d}(\mathbf{p})].
    \label{eq:moho-data-misfit}
\end{equation}

Function $\phi(\mathbf{p})$ is non-linear with respect to $\mathbf{p}$.
Thus, we can determine its minimum using gradient-based
iterative optimization
methods like Gauss-Newton or Steepest Descent.
Such methods start from an initial approximation to the model parameter vector
$\mathbf{p}^0$ and estimate a parameter perturbation vector
$\mathbf{\Delta p}^0$.
The perturbation vector is used to update $\mathbf{p}^0$ to
$\mathbf{p}^1 =  \mathbf{p}^0 + \mathbf{\Delta p}^0$.
This procedure is repeated until a minimum of function $\phi(\mathbf{p})$
(Eq.~\ref{eq:moho-data-misfit}) is reached.

For the Gauss-Newton method,
the parameter perturbation vector at the $k$th iteration $\mathbf{\Delta p}^k$
is obtained by solving the linear system

\begin{equation}
    \mathbf{H}^k\mathbf{\Delta p}^k = -\mathbf{\nabla\phi}^k,
    \label{eq:moho-gaussnewton}
\end{equation}

@@

@@

\begin{equation}
    \mathbf{\nabla\phi}^k = -2{\mathbf{A}^k}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)],
    \label{eq:moho-gradient}
\end{equation}

@@

\begin{equation}
    \mathbf{H}^k \approx 2{\mathbf{A}^k}^T{\mathbf{A}^k},
    \label{eq:moho-hessian}
\end{equation}

\noindent in which
$\mathbf{A}^k$ is the $N \times M$ Jacobian or sensitivity matrix
whose elements are

\begin{equation}
    A_{ij}^k = \dfrac{\partial f_i}{\partial p_j}(\mathbf{p}^k).
    \label{eq:moho-jacobian}
\end{equation}

@@

Non-linear gravity inversions for estimating the relief of an interface
separating two media (like the Moho) are ill-posed and require additional
constraints in the form of regularization \citep{silva2001b}.
A common approach is to use the first-order Tikhonov regularization
\citep{tikhonov1977} to impose smoothness on the solution.
The cost function for smoothness regularization is given by

\begin{equation}
    \theta(\mathbf{p}) = \mathbf{p}^T\mathbf{R}^T\mathbf{R}\mathbf{p},
    \label{eq:moho-regul}
\end{equation}

\noindent where $\mathbf{R}$ is an $L \times M$ finite-difference matrix
representing $L$ first-order differences between adjacent tesseroids.

To transform the ill-posed inverse problem into a well-posed one via Tikhonov
regularization, we adopted the well-established procedure
of formulating a constrained inverse problem that is solved by minimizing an
unconstrained goal function

\begin{equation}
    \Gamma(\mathbf{p}) = \phi(\mathbf{p}) + \mu\theta(\mathbf{p}),
    \label{eq:moho-goalfunction}
\end{equation}

\noindent
in which $\mu$ is the regularization parameter that controls the balance
between fitting the observed data and obeying the smoothness constraint imposed
by the regularizing function $\theta(\mathbf{p})$ (Eq.~\ref{eq:moho-regul}).

The goal function $\Gamma(\mathbf{p})$ is also non-linear with respect to
$\mathbf{p}$ and can be minimized using the Gauss-Newton method.
The gradient vector and Hessian matrix of the goal function are, respectively,

\begin{equation}
    \mathbf{\nabla\Gamma}^k =
        -2{\mathbf{A}^k}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)] +
        2\mu\mathbf{R}^T\mathbf{R}\mathbf{p}^k,
    \label{eq:moho-gradient-regul}
\end{equation}

@@

\begin{equation}
    \mathbf{H}^k = 2{\mathbf{A}^k}^T{\mathbf{A}^k}
                   + 2\mu\mathbf{R}^T\mathbf{R}.
    \label{eq:moho-hessian-regul}
\end{equation}

At the $k$th iteration, the parameter perturbation vector
$\mathbf{\Delta p}^k$ is obtained by solving the linear equation system

\begin{equation}
    \left[{\mathbf{A}^k}^T{\mathbf{A}^k} + \mu\mathbf{R}^T\mathbf{R}\right]
    \mathbf{\Delta p}^k =
    {\mathbf{A}^k}^T[\mathbf{d}^o - \mathbf{d}(\mathbf{p}^k)] -
    \mu\mathbf{R}^T\mathbf{R}\mathbf{p}^k.
    \label{eq:moho-gaussnewton-regul}
\end{equation}

Estimating the Moho depths using the above equations is computationally
costly because of two main factors:
(1) the evaluation and storage of the dense $N \times M$ Jacobian matrix
${\mathbf{A}^k}$
and (2) the solution of the resulting $M \times M$ equation system
(not required for Steepest Descent).
In practice, the derivatives in the Jacobian (Eq.~\ref{eq:moho-jacobian})
are often calculated through a first-order finite-difference approximation.
Thus, evaluating ${\mathbf{A}^k}$ requires $2\times M \times N$ forward modeling
operations for each iteration of the gradient descent algorithm.
These computations are performed for each iteration of the optimization
of the goal function $\Gamma(\mathbf{p})$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software implementation}\label{sec:software}
